Comparing OTU quality across clustering strategies

### 2022-01-13 paper/response-to-reviewers.md

**The manuscript is well written and describes a new computational tool
without computer science jargon. I think this manuscript and the method
will be very well received because it does fill a gap in the computation
analysis of amplicon sequences. A have only a few comments and
suggestions as outlined below:**

Thank you for your enthusiasm!

**OTU quality: The main tool for comparing OTU quality in this
manuscript is MCC. This metric is central to de novo OTU clustering in
the OptiClust approach, and I think it greatly benefits amplicon-based
approaches. The MCC metric is briefly described on Lines 54-55. However,
since it is so central to the comparisons in the current study, I wonder
if the reader would benefit from a bit more detail and, I think the
reader needs to be reminded about MCC when interpreting the results.
When reviewing Fig 1: the reader should know exactly why an MCC of 0.91
is better than 0.78 and have a way to conceptualizing the difference.
Also, it may be helpful to remind the reader how quality was determined
(see L172).**

We describe MCC in more detail under the description of the OptiFit
algorithm, which we expect readers would read just before looking at
Figure 1. We added a few more details to the MCC description in this
section (L111-113 and L117-118), In the caption of Figure 1 we also
added examples of the confusion matrix values for the first and last MCC
scores.

We revised the sentence formerly at L172 to note that we mean “quality”
in terms of MCC score (L184).

@@

**OptiClust: On lines 78-80 the authors state that “previous studies”
found OptiClust generate the highest quality OTUs, but then they have
only a single reference which is the original paper describing the
method (and from the corresponding authors research group). Full
disclosure, I use OptiClust. However, I think the authors need to either
say “A previous study” or provide additional references (and perhaps one
from outside the research group).**

Thank you for catching this. We have revised this sentence to say “We
previously found” in order to clarify that this claim is supported by a
single study from our group (L86).

@@

**Computational Resources: Throughout the manuscript, the authors often
use compute time as a measure of the computational resource demand. This
is an important metric, but I also wonder about RAM. Specifically, at
the end of the manuscript the authors state “we recommend using
OptiClust for de novo sequencing over the split strategy with OptiFit
since OptiClust is simpler to execute and performs similarly in terms of
both run time and OTU quality” (L279-281). Earlier they also stated that
OptiClust was faster (L244 and L257). Regardless of time comparisons,
I’m also wondering if the OptiFit split strategy could be an added
benefit for individuals that lack computation performance (e.g., RAM). I
don’t remember from the OptiClust paper how much RAM is needed or if it
scaled-up with data size, but I thought it was interesting that 10% of a
sequencing run could be used to generate de novo sequences with
OptiClust and then OptiFit could be used to match to other 90% without
sacrificing quality. Could this be a benefit for research groups that
may be computationally limited?**

Saving computational resources is an interesting thought! However, we
found that the RAM usage was similar between OptiClust and OptiFit
regardless of the reference fraction size used in the split strategy.
Both OptiClust and OptiFit consider the distances between all pairs of
sequences, so changing the reference fraction and using the remaining
sequences as the query doesn’t change the total number of distances that
are considered in the MCC. We have added statements to note how the RAM
usage of OptiFit & OptiClust compared (L217-219) and how the RAM usage
compared across reference fraction sizes for the split dataset strategy
(L233-235). We chose not to include the RAM data in the figures since we
found RAM to be similar, and the figures already contain a lot of
information.

@@

**Other Reference Based Methods: I think the authors make a good point
when the compare VSEARCH and OptiFit (L261-271). Specifically, I found
it interesting to think about how VSEARCH can allow more dissimilar
sequences to be clustered into the reference OTU. I think this is
something that is worth turning into a stronger statement. For example,
since VSEARCH could cluster 3% dissimilar to a centroid without
considering the other sequences being clustered, does that mean a
maximum of 6% dissimilar across all reference clustered sequences is
theoretically possible? In other words, is VSEARCH clustering more in
the closed approach because the within-OTU maximum dissimilarity is
actually very high? This was just a though, but perhaps another way to
strengthen the discussion.**

We agree, it is theoretically possible that when using VSEARCH for
reference clustering, query sequences may be 3% dissimilar to a centroid
sequence but possibly up to 6% dissimilar to each other. Since we did
not investigate this directly, we hesitate to make *too* strong of a
statement in the discussion, but we did add a statement (L289) to better
emphasize this point.

@@

## Reviewer #2 (Comments for the Author):

**The manuscript introduces optiFit, an algorithm that builds on
OptiClust that aims to provide an efficient and robust method to fit
amplicon data to existing OTUs. The authors utilize OptiClust output as
their starting list of reference OTUs, and aim to cluster their new
query sequences to the existing output. The results section is logical
and very well written: It describes how the algorithm works, the
benchmarks its performance with various reference databases (SILVA,
GreenGenes, and RDP), compare the results with QIME2 clustering
approaches.**

**The reviewer is by no means an expert in computational biology, and
merely has a microbial ecology based background. As such, issues
regarding performance metrics, algorithm strategy, coefficient, etc..
are beyond my expertise. Judging merely based on the need for such
algorithm, the overall strategy, and the apparent results and
benchmarking using reference datasets and databases, I find the results
convincing. The need for such addition to the suite of approaches
offered by MOTHUR is also well explained. I encourage the authors to
define what “high quality OTUs” and “low quality OTUs” means and how
they are assessed early on in the introduction.**

Thank you for your feedback!

We added a sentence in the introduction to clarify the meaning of
quality (L54-57), and provided additional details of the MCC score under
the description of the OptiFit algorithm (L111-113 and L117-118).

@@

**The OptiFit algorithm is a natural extension of the OptiClust approach
to sequence classification, albeit for situations where a
reference-based approach is either required or beneficial. One of the
more appealing aspects of OptiFit is that it does not change OTU
assignments for samples/data that have been already analyzed when adding
new information to a dataset. This functionality would be particularly
useful for investigators who are carrying out meta-analysis type studies
or who have very large datasets to which new samples are added regularly
and reanalysis would be computationally burdensome.**

**A general question I have is whether the authors have considered
suggesting that users make use of the data-splitting approach that they
outline in their own paper. OTU classification is a duel task problem
(identifying OTUs in a dataset and assigning each read sequence to an
appropriate OTU). With de novo classification approaches the same data
is used to both identify and classify the reads. From my perspective,
the development of OptiFit makes a perfect partner for OptiClust in a
Train/Test (Train/Classify) splitting situation, where the users would
split data, train using OptiClust and then classify the remaining reads
using OptiFit. Of course, pursuing this would be well beyond the scope
of this manuscript, but the authors may consider discussing such a
possibility for future work.**

We are pleased that you made this connection! We agree that it is well
beyond the scope of this manuscript, and we are already working on this
for another forthcoming paper. Thus, we chose to merely mention machine
learning alongside the other potential applications in order to keep
this paper narrowly focused on the OptiFit algorithm.

@@

**The manuscript is generally clear, concise, and well written. There
are only so many times that one can write “clustered”/“clustering” (or
other variants) in a sentence or paragraph and still have it make sense.
I can tell (and appreciate) that the authors took great care in the
writing process to properly convey their message.**

Thank you!

**Specific comments:**

Thank you for your thorough reading of the manuscript and the detailed
comments you provided below.

@@

**Line 4 (Abstract): The sentence discussing OptiClust appears to be
tangential to the opening of the abstract. Perhaps it would be better to
add this context further down in the abstract (line 16-ish).**

We have moved the mention of OptiClust near where you suggested (L17).
We also trimmed down the abstract for brevity.

@@

**Line 83-85: “While other tools represent reference OTUs with a single
sequence, OptiFit 84 uses multiple sequences in existing OTUs as the
reference and fits new sequences to 85 those reference OTUs.” Should
“multiple” be “all” in this sentence?**

Yes, thank you! (L92)

**Lines 129 - 131: Please include references for Greengenes and SILVA,
and indicate what versions of all three (Greengenes, SILVA, and RDP)
were used, even though this information is included in the Methods
section.**

References to all three were included at the end of the sentence in the
order they were mentioned. To clear up any confusion, we moved each
database’s reference to immediately after its mention, and also noted
the version numbers here as you suggested (L141-143).

@@

**Lines 141-142: “Thus, open reference OptiFit produced OTUs of very
similar quality as de novo clustering, …” please add “with OptiClust” to
be more specific about what the comparison being made is.**

Agreed. (L155)

**Figure 1: W-Z are designated as the query sequences in the caption. Z
does not appear in the distance threshold table at the top of the figure
and is not assigned to an OTU at the end of Cycle 4. From Line 115
(“Alternatively, a sequence will remained unassigned…”) the reader can
infer that sequence Z is different enough from the others that it was
designated a singleton (outside of the distance threshold discussed in
lines 111-113). If I’m correct, I encourage the authors to make this
point explicitly in Figure 1’s caption.**

That’s correct, we added a sentence to the caption as you suggested.

@@

**Figure 3: I suggest changing the label for “de novo” to “OptiClust (de
novo)” or similar to emphasize that OptiClust was being used for those
benchmark tests.**

In Figure 3, we show results from *de novo* clustering with both the
OptiClust algorithm in mothur (red triangles) and VSEARCH (blue
triangles). Changing the label to “OptiClust (*de novo*)” would not be
accurate, as VSEARCH does not use the OptiClust algorithm. Instead, we
changed the red color label from “mothur” to “OptiClust (\_de novo) or
OptiFit” in Figure 3.

### 2022-01-14 paper/response-to-reviewers.Rmd

**The manuscript is well written and describes a new computational tool without
computer science jargon. I think this manuscript and the method will be very
well received because it does fill a gap in the computation analysis of amplicon
sequences. A have only a few comments and suggestions as outlined below:**

Thank you for your enthusiasm!

**OTU quality: The main tool for comparing OTU quality in this manuscript is
MCC. This metric is central to de novo OTU clustering in the OptiClust approach,
and I think it greatly benefits amplicon-based approaches. The MCC metric is
briefly described on Lines 54-55. However, since it is so central to the
comparisons in the current study, I wonder if the reader would benefit from a
bit more detail and, I think the reader needs to be reminded about MCC when
interpreting the results. When reviewing Fig 1: the reader should know exactly
why an MCC of 0.91 is better than 0.78 and have a way to conceptualizing the
difference. Also, it may be helpful to remind the reader how quality was
determined (see L172).**

We describe MCC in more detail under the description of the OptiFit algorithm, 
which we expect readers would read just before looking at Figure 1.
We added a few more details to the MCC description in this section (L111-113 and
L117-118), 
In the caption of Figure 1 we also added examples of the confusion matrix 
values for the first and last MCC scores.

We revised the sentence formerly at L172 to note that we mean "quality" in terms
of MCC score (L184).

@@

**OptiClust: On lines 78-80 the authors state that "previous studies" found
OptiClust generate the highest quality OTUs, but then they have only a single
reference which is the original paper describing the method (and from the
corresponding authors research group). Full disclosure, I use OptiClust.
However, I think the authors need to either say "A previous study" or provide
additional references (and perhaps one from outside the research group).**

Thank you for catching this.
We have revised this sentence to say "We previously found" in order to clarify
that this claim is supported by a single study from our group (L86).

@@

**Computational Resources: Throughout the manuscript, the authors often use
compute time as a measure of the computational resource demand. This is an
important metric, but I also wonder about RAM. Specifically, at the end of the
manuscript the authors state "we recommend using OptiClust for de novo
sequencing over the split strategy with OptiFit since OptiClust is simpler to
execute and performs similarly in terms of both run time and OTU quality"
(L279-281). Earlier they also stated that OptiClust was faster (L244 and L257).
Regardless of time comparisons, I'm also wondering if the OptiFit split strategy
could be an added benefit for individuals that lack computation performance
(e.g., RAM). I don't remember from the OptiClust paper how much RAM is needed or
if it scaled-up with data size, but I thought it was interesting that 10% of a
sequencing run could be used to generate de novo sequences with OptiClust and
then OptiFit could be used to match to other 90% without sacrificing quality.
Could this be a benefit for research groups that may be computationally
limited?**

Saving computational resources is an interesting thought! 
However, we found that the RAM usage was similar between OptiClust and OptiFit
regardless of the reference fraction size used in the split strategy.
Both OptiClust and OptiFit consider the distances between all pairs of
sequences, so changing the reference fraction and using the remaining sequences
as the query doesn't change the total number of distances that are considered in
the MCC.
We have added statements to note how the RAM usage of OptiFit & OptiClust
compared (L217-219) and how the RAM usage compared across reference fraction
sizes for the split dataset strategy (L233-235).
We chose not to include the RAM data in the figures since we found RAM to be
similar, and the figures already contain a lot of information.

@@

**Line 83-85: "While other tools represent reference OTUs with a single
sequence, OptiFit 84 uses multiple sequences in existing OTUs as the reference
and fits new sequences to 85 those reference OTUs." Should "multiple" be "all"
in this sentence?**

Yes, thank you! (L92)

**Lines 129 - 131: Please include references for Greengenes and SILVA, and
indicate what versions of all three (Greengenes, SILVA, and RDP) were used, even
though this information is included in the Methods section.**

References to all three were included at the end of the sentence in the order
they were mentioned.
To clear up any confusion, we moved each database's reference to immediately
after its mention, and also noted the version numbers here as you suggested (L141-143).

@@

**Lines 141-142: "Thus, open reference OptiFit produced OTUs of very similar
quality as de novo clustering, ..." please add "with OptiClust" to be more
specific about what the comparison being made is.**

Agreed. (L155)

**Figure 1: W-Z are designated as the query sequences in the caption. Z does not
appear in the distance threshold table at the top of the figure and is not
assigned to an OTU at the end of Cycle 4. From Line 115 ("Alternatively, a
sequence will remained unassigned...") the reader can infer that sequence Z is
different enough from the others that it was designated a singleton (outside of
the distance threshold discussed in lines 111-113). If I'm correct, I encourage
the authors to make this point explicitly in Figure 1's caption.**

That's correct, we added a sentence to the caption as you suggested.

@@

**Figure 3: I suggest changing the label for "de novo" to "OptiClust (de novo)"
or similar to emphasize that OptiClust was being used for those benchmark
tests.**

In Figure 3, we show results from _de novo_ clustering with both the OptiClust
algorithm in mothur (red triangles) and VSEARCH (blue triangles). 
Changing the label to "OptiClust (_de novo_)" would not be accurate, as VSEARCH 
does not use the OptiClust algorithm.
Instead, we changed the red color label from "mothur" to "OptiClust (_de novo)
or OptiFit" in Figure 3.

### 2022-01-14 exploratory/2022/sensspec_2022-01.Rmd

```{r conf_matrix, fig.dim=c(6.87,6.87)}
dat_agg %>% 
  filter(dataset == 'human', 
         ref == 'gg' | strategy != 'database') %>%
  pivot_longer(c(tp, tn, fp, fn), 
               names_to = 'conf_mat_quad', 
               values_to = 'conf_mat_val') %>%
  mutate(conf_mat_quad = factor(conf_mat_quad, 
                                levels = c('tp', 'fn', 'fp', 'tn'))) %>% 
  # mutate(tool_method = factor(glue("{tool}_{method}"),
  #                             levels = c("mothur_de_novo", "vsearch_de_novo",
  #                                        "mothur_closed", "vsearch_closed",
  #                                        "mothur_open", "vsearch_open"))
  #        ) %>% 
  ggplot(aes(
    strategy,
    conf_mat_val,
    color = tool,
    shape = method
  )) +
  facet_wrap("conf_mat_quad", 
             scales = 'free_y',
             nrow = 2) +
  stat_summary(
    geom = "point",
    fun = median,
    size = 2,
    position = position_dodge(width = 0.4)
  ) +
  scale_shape_manual(values = list(
    open = 1,
    closed = 19,
    de_novo = 17
  )) +
  labs(y = '', x = '', title = 'Confusion Matrix for the human dataset')
```

>>>>

```{r conf_matrix, fig.dim=c(6.87,5)}
dat_agg %>% 
  filter(dataset == 'human', 
         ref == 'gg' | strategy != 'database') %>%
  pivot_longer(c(tp, tn, fp, fn), 
               names_to = 'conf_mat_quad', 
               values_to = 'conf_mat_val') %>%
  mutate(conf_mat_quad = factor(conf_mat_quad, 
                                levels = c('tp', 'fn', 'fp', 'tn'))) %>% 
  # mutate(tool_method = factor(glue("{tool}_{method}"),
  #                             levels = c("mothur_de_novo", "vsearch_de_novo",
  #                                        "mothur_closed", "vsearch_closed",
  #                                        "mothur_open", "vsearch_open"))
  #        ) %>% 
  ggplot(aes(
    strategy,
    conf_mat_val,
    color = tool,
    shape = method
  )) +
  facet_wrap("conf_mat_quad", 
             scales = 'free_y',
             nrow = 2) +
  stat_summary(
    geom = "point",
    fun = median,
    size = 2,
    position = position_dodge(width = 0.4)
  ) +
  scale_shape_manual(values = list(
    open = 1,
    closed = 19,
    de_novo = 17
  )) +
  labs(y = '', x = '', title = 'Confusion Matrix for the human dataset')
```
